---
title: "Filter genotypes, wecare feb2016"
author: "Alexey Larionov"
output: html_document
---

started: 01Mar2016
last updated: 12May2016

# Notes 

It was discussed with DC whether to filter cases by call rate per case. 
There was ~3 cases with low coverage (<20) and low call rates (<50%). 
We desided to keep such cases because their retained genotypes still passed all filters. 

Instead of setting the thresholds by the proportions of reads (as done in this script), 
it is possible to set thresholds by probability of the observed reads. 
This could be dome using pbinom(count of minor reads, sum of reads, 0.5). 
The used proportion thresholds correspond to 1-5%, depending on total reads count. 

Thresholds for min DP and GQ were set to 10 and 20 because they are commonly used 
(e.g. see Carson BMC Bioinformatics 2014 15:125). 

A small number of genotypes (~0.01%) was covered too high to be true (up to 1-2k coverage).
These are obvious mistakes, and they have been removed too.  Arbitrary the threshold for
max DP was set to 500 (appr. 10 fold of average coverage).

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Clean-up
rm(list=ls())
graphics.off()

# Set root working folder
library(knitr)

opts_knit$set(root.dir = "/scratch/medgen/scripts/rscripts_05.16")
#setwd("/scratch/medgen/scripts/rscripts_05.16")

#opts_knit$set(root.dir = "C:\\Users\\larion01\\Documents\\GitHub\\R1")
#setwd("C:\\Users\\larion01\\Documents\\GitHub\\R1")

```

# filtering_settings

```{r filtering_settings}

# Set thresholds for cumulative reads depth and genotype quality
min.dp <- 10
max.dp <- 500
min.gq <- 20

# Minimal call rate per variant
min.call.rate <- 0.8

# Set theresholds for fraction of reads with minor allele (frma)
max.frma.hom <- 0.05 # Max frma for homozigous calls
min.frma.het <- 0.35 # Min frma for heterozigous calls

```

# load_and_check_data

```{r load_and_check_data}

load(file="data/s02_filter_variants_feb2016.RData")
ls()

dim(covar.df)
str(covar.df)
covar.df[1:5,1:5]

dim(demographics.df)
str(demographics.df)
demographics.df[1:5,1:5]

dim(samples.df)
str(samples.df)
samples.df[1:5,]

dim(vv.pf.df)
str(vv.pf.df)
vv.pf.df[1:5,1:5]

dim(gt.pf.df)
str(gt.pf.df, list.len=5)
gt.pf.df[1:5,1:5]

dim(gq.pf.df)
str(gq.pf.df, list.len=5)
gq.pf.df[1:5,1:5]

dim(dp.pf.df)
str(dp.pf.df, list.len=5)
dp.pf.df[1:5,1:5]

dim(ref.pf.mx)
class(ref.pf.mx)
ref.pf.mx[1:15,1:5]

dim(alt.pf.mx)
class(alt.pf.mx)
alt.pf.mx[1:15,1:5]

# Check consistence of rownames
sum(rownames(gt.pf.df) != rownames(vv.pf.df), na.rm=TRUE)

sum(rownames(gt.pf.df) != rownames(gq.pf.df), na.rm=TRUE)
sum(rownames(gt.pf.df) != rownames(dp.pf.df), na.rm=TRUE)
sum(rownames(gt.pf.df) != rownames(ref.pf.mx), na.rm=TRUE)
sum(rownames(gt.pf.df) != rownames(alt.pf.mx), na.rm=TRUE)

# Check consistence of colnames (sub(".DP","") may be used instead)
sum(substr(colnames(gt.pf.df),0,6) != substr(colnames(gq.pf.df),0,6), na.rm=TRUE)
sum(substr(colnames(gt.pf.df),0,6) != substr(colnames(dp.pf.df),0,6), na.rm=TRUE)
sum(substr(colnames(gt.pf.df),0,6) != substr(colnames(ref.pf.mx),0,6), na.rm=TRUE)
sum(substr(colnames(gt.pf.df),0,6) != substr(colnames(alt.pf.mx),0,6), na.rm=TRUE)

```

# convert_data_frames_to_matrices

```{r convert_data_frames_to_matrices}

gt.mx <- as.matrix(gt.pf.df)
gq.mx <- as.matrix(gq.pf.df)
dp.mx <- as.matrix(dp.pf.df)
rm(gt.pf.df, gq.pf.df, dp.pf.df)

dim(gt.mx)
class(gt.mx)
gt.mx[1:5,1:5]

dim(gq.mx)
class(gq.mx)
gq.mx[1:5,1:5]

dim(dp.mx)
class(dp.mx)
dp.mx[1:5,1:5]

```

# convert_ref_alt_matrices_to_numeric

```{r }

ref.alt.rownames <- rownames(ref.pf.mx)
ref.alt.colnames <- colnames(ref.pf.mx)

ref.mx <- matrix(as.numeric(ref.pf.mx), nrow=nrow(ref.pf.mx))
rownames(ref.mx) <- ref.alt.rownames
colnames(ref.mx) <- ref.alt.colnames

dim(ref.mx)
class(ref.mx)
ref.mx[1:15,1:5]

alt.mx <- matrix(as.numeric(alt.pf.mx), nrow=nrow(alt.pf.mx))
rownames(alt.mx) <- ref.alt.rownames
colnames(alt.mx) <- ref.alt.colnames

dim(alt.mx)
class(alt.mx)
alt.mx[1:15,1:5]

rm(ref.alt.rownames, ref.alt.colnames, ref.pf.mx, alt.pf.mx)

```

# histograms_before_filtering

```{r histograms_before_filtering}

# Total num of NAs before genotype filtering
sum(is.na(gt.mx))/(dim(gt.mx)[1]*dim(gt.mx)[2]) # <1%

# Histograms of all gq and dp
hist(gq.mx, breaks=50, main="Histograms of all gq", xlab=NULL)
hist(gq.mx, breaks=250, xlim=c(80,100), main="Histograms of all gq (zoom to 80-100)", xlab=NULL)

hist(dp.mx, breaks=50, main="Histograms of all dp", xlab=NULL)
hist(dp.mx, breaks=500, xlim=c(0,150), main="Histograms of all dp (zoom to 0-150)", xlab=NULL)

# Histograms of gq and dp in retained genotypes
hist(gq.mx[!is.na(gt.mx)], breaks=50, main="Histograms of gq before filtering (gt is not NA)", xlab=NULL)
hist(gq.mx[!is.na(gt.mx)], breaks=250, xlim=c(80,100), main="Histograms of gq before filtering (gt is not NA, zoom to 80-100)", xlab=NULL)

hist(dp.mx[!is.na(gt.mx)], breaks=50, main="Histograms of dp before filtering (gt is not NA)", xlab=NULL)
hist(dp.mx[!is.na(gt.mx)], breaks=500, xlim=c(0,150), main="Histograms of dp before filtering (gt is not NA, zoom to 0-150)", xlab=NULL)

# Call rates per variant before genotype filtering
x <- ncol(gt.mx)
y <- apply(gt.mx,1,function(z){1-sum(is.na(z))/x})
hist(y, breaks=50, xlab=NULL, main="Call rates per variant before genotypes filtering")
hist(y, breaks=500, xlim=c(0.9,1), xlab=NULL, main="Call rates per variant before genotypes filtering (zoom 0.9-1)")
rm(x,y)

```

# filter_by_gq

```{r filter_by_gq}

sum(gq.mx < min.gq, na.rm=TRUE)
sum(gq.mx < min.gq, na.rm=TRUE)/sum(!is.na(gq.mx))

NA -> gt.mx[ gq.mx < min.gq ]
rm(min.gq)

# num of NAs after filter
sum(is.na(gt.mx))/(dim(gt.mx)[1]*dim(gt.mx)[2]) # ~5%

```

# filter_by_dp

```{r filter_by_dp}

sum(dp.mx < min.dp, na.rm=TRUE)
sum(dp.mx < min.dp, na.rm=TRUE)/sum(!is.na(gq.mx))

sum(dp.mx > max.dp, na.rm=TRUE)
sum(dp.mx > max.dp, na.rm=TRUE)/sum(!is.na(gq.mx))

NA -> gt.mx[ dp.mx < min.dp ]
NA -> gt.mx[ dp.mx > max.dp ]

rm(min.dp, max.dp)

# num of NAs after filter
sum(is.na(gt.mx))/(dim(gt.mx)[1]*dim(gt.mx)[2]) # ~7%

```

# calculate_fractions_of_reads_with_minor_allele

```{r calculate_fractions_of_reads_with_minor_allele}

# Sum of ref and alt
sum.mx <- ref.mx + alt.mx

dim(sum.mx)
sum.mx[1:15,1:5]

# Subtract alt from ref
ref.sub.alt <- ref.mx - alt.mx

dim(ref.sub.alt)
ref.sub.alt[1:15,1:5]

# Count reads with minor allele (crma)
crma.mx <- ref.sub.alt
crma.mx[ref.sub.alt == 0] <- ref.mx[ref.sub.alt == 0] # use any (e.g. ref) if no difference
crma.mx[ref.sub.alt < 0] <- ref.mx[ref.sub.alt < 0] # use ref, if itis smaller
crma.mx[ref.sub.alt > 0] <- alt.mx[ref.sub.alt > 0] # use alt, if it is smaller

dim(crma.mx)
crma.mx[1:15,1:5]

ref.mx[1:15,1:5]
alt.mx[1:15,1:5]

rm(ref.sub.alt)

# Calculate fractions of reads with minor allele (frma)
frma.mx <- crma.mx/sum.mx

dim(frma.mx)
frma.mx[1:15,1:5]

crma.mx[1:15,1:5]
sum.mx[1:15,1:5]

# These checks are identical
# NA is any Non Available data (e.g. something was missed)
# NaN = Not a Number (e.g. Infinity, when genotypes with zero sum of ref + alt ...)?
sum(is.na(frma.mx)) 
sum(is.nan(frma.mx)) 
sum(!is.finite(frma.mx)) 

rm(crma.mx, sum.mx)

```

# filter_by_fraction_of_reads_with_minor_allele

```{r filter_by_fraction_of_reads_with_minor_allele}

# num of NAs before filters
sum(is.na(gt.mx))
sum(is.na(gt.mx))/(dim(gt.mx)[1]*dim(gt.mx)[2]) # ~7%

# Filter out genotype calls with uncertain maf.
# This picks a very small number of genotypes (20 of 10M), which somehow had been
# called and passed DP and GQ filters despite both Ref and Alt being Zero ...
# It's scary to have such genotype calls ...
NA -> gt.mx[ !is.finite(frma.mx) ]

# num of NAs after filter
sum(is.na(gt.mx))
sum(is.na(gt.mx))/(dim(gt.mx)[1]*dim(gt.mx)[2]) # ~7%

# Filter homozigous genotype calls (i.e. coded 0 or 2)
# that have too high minor allele reads fraction
NA -> gt.mx[ gt.mx != 1 & frma.mx > max.frma.hom ]
rm(max.frma.hom)

# num of NAs after filter
sum(is.na(gt.mx))
sum(is.na(gt.mx))/(dim(gt.mx)[1]*dim(gt.mx)[2]) # ~7%

# Filter heterozigous genotype calls (i.e. coded 1)
# that have too low minor allele reads fraction
NA -> gt.mx[ gt.mx == 1 & frma.mx < min.frma.het ]
rm(frma.mx, min.frma.het)

# num of NAs after filter
sum(is.na(gt.mx))
sum(is.na(gt.mx))/(dim(gt.mx)[1]*dim(gt.mx)[2]) # ~8%

```

# filter_variants_by_call_rate

```{r filter_variants_by_call_rate}

# Call rates per variant before after genotypes filtering
x <- ncol(gt.mx)
y <- apply(gt.mx, 1, function(z){1-sum(is.na(z))/x})
hist(y, breaks=100, xlab=NULL, main="Call rates per variant after genotypes filtering")

# Set filter and estimate the proportion of variants to remove
var.retained <- y >= min.call.rate
sum(var.retained)
1-sum(var.retained)/nrow(gt.mx)
rm(x,y,min.call.rate)

# Remove variants with loaw call rates
gt.mx <- gt.mx[ var.retained, ]
gq.mx <- gq.mx[ var.retained, ]
dp.mx <- dp.mx[ var.retained, ]
alt.mx <- alt.mx[ var.retained, ]
ref.mx <- ref.mx[ var.retained, ]
vv.df <- vv.pf.df[ var.retained, ]
rm(var.retained, vv.pf.df)

# Call rates per variant after filtering by call rate
x <- ncol(gt.mx)
y <- apply(gt.mx, 1, function(z){1-sum(is.na(z))/x})
hist(y, breaks=20, xlab=NULL, xlim=c(0,1), main="Call rates per variant after filtering by call rate")
rm(x,y)

```

# histograms_after_filtering

```{r histograms_after_filtering}

# Total num of NAs after genotypes filtering and removal of variants with low call rate
sum(is.na(gt.mx))/(dim(gt.mx)[1]*dim(gt.mx)[2]) # ~4%

# Histograms of gq and dp after filtering
hist(gq.mx[!is.na(gt.mx)], breaks=50, xlim=c(0,100), main="Histograms of gq after filtering", xlab=NULL)
hist(gq.mx[!is.na(gt.mx)], breaks=200, xlim=c(80,100), main="Histograms of gq after filtering (zoom to 80-100)", xlab=NULL)

hist(dp.mx[!is.na(gt.mx)], breaks=50, main="Histograms of dp after filtering", xlab=NULL)
hist(dp.mx[!is.na(gt.mx)], breaks=150, xlim=c(0,150), main="Histograms of dp after filtering (gt is not NA, zoom to 0-150)", xlab=NULL)

# Call rates per sample: based on variants retained after all filters
x <- nrow(gt.mx)
y <- apply(gt.mx, 2, function(z){1-sum(is.na(z))/x})
barplot(y, ylab="Callrate", ylim=c(0,1), xlab="Samples", xaxt="n", 
  main="Call rates per sample: based on variants retained after all filters")
rm(x,y)

```

# save_data

```{r save_data}

save.image(file="data/s03_filter_genotypes_feb2016.RData")

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```

