---
title: "Simulate data and explore rare varints tests"
author: "Alexey Larionov"
output: html_document
---

started: 24Apr2016
last updated: 24Apr2016

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Clean-up
rm(list=ls())
graphics.off()

# Set root working folder
library(knitr)
opts_knit$set(root.dir = "/scratch/medgen/users/alexey/wecare_feb2016/wecare_feb2016_r")
#setwd("/scratch/medgen/users/alexey/wecare_feb2016/wecare_feb2016_r")

```

# simulate_data

```{r simulate_data}

# Outcomes: 100 cases (1) and 100 controls (0)
Y <- c(rep(1,100),rep(0,100))

# Genotypes matrix
G <- matrix(rep(0,1400), nrow=7)

# Variant 1: common, risk [20 of 100] vs [10 of 100]
G[1,1:20] <- rep(1,20)
G[1,101:110]  <- rep(1,10)

# Variant 2: rare, risk [2 of 100] vs [1 of 100]
G[2,1:2] <- rep(1,2)
G[2,101]  <- 1

# Variant 3: common, protective [10 of 100] vs [20 of 100]
G[3,1:10] <- rep(1,10)
G[3,101:120]  <- rep(1,20)

# Variant 4: rare, protective [1 of 100] vs [2 of 100]
G[4,1] <- 1
G[4,101:102]  <- rep(1,2)

# Variant 5: rare, no association [2 of 100] vs [2 of 100]
G[5,1:2] <- rep(1,2)
G[5,101:102]  <- rep(1,2)

# Variant 6: common, no association [20 of 100] vs [20 of 100]
G[6,1:20] <- rep(1,20)
G[6,101:120]  <- rep(1,20)

# Variant 7: common, near perfect association [99 of 100] vs [1 of 100]
G[7,1:99] <- rep(1,99)
G[7,101]  <- 1

# Covariates matrix
#D <- 

```

# explore_model_for_Variant_1
common, risk [20 of 100] vs [10 of 100]

```{r explore_model_for_Variant_1}

Var1 <- G[1,]
M1 <-glm(Y ~ Var1, family = binomial(link = "logit")) 
summary(M1)
coef(summary(M1))["Var1", "Estimate"]
coef(summary(M1))["Var1", "Std. Error"]
coef(summary(M1))["Var1", "z value"]
coef(summary(M1))["Var1", "Pr(>|z|)"]
logLik(M1)
rm(Var1)

```

# explore_model_for_Variant_2
rare, risk [2 of 100] vs [1 of 100]

```{r explore_model_for_Variant_2}

Var2 <- G[2,]
M2 <-glm(Y ~ Var2, family = binomial(link = "logit")) 
summary(M2)
coef(summary(M2))["Var2", "Estimate"]
coef(summary(M2))["Var2", "Std. Error"]
coef(summary(M2))["Var2", "z value"]
coef(summary(M2))["Var2", "Pr(>|z|)"]
logLik(M2)
rm(Var2)

```

# explore_model_for_Variant_3
common, protective [10 of 100] vs [20 of 100]

```{r explore_model_for_Variant_3}

Var3 <- G[3,]
M3 <-glm(Y ~ Var3, family = binomial(link = "logit")) 
summary(M3)
coef(summary(M3))["Var3", "Estimate"]
coef(summary(M3))["Var3", "Std. Error"]
coef(summary(M3))["Var3", "z value"]
coef(summary(M3))["Var3", "Pr(>|z|)"]
logLik(M3)
rm(Var3)

```

# explore_model_for_Variant_4
rare, protective [1 of 100] vs [2 of 100]

```{r explore_model_for_Variant_4}

Var4 <- G[4,]
M4 <-glm(Y ~ Var4, family = binomial(link = "logit")) 
summary(M4)
coef(summary(M4))["Var4", "Estimate"]
coef(summary(M4))["Var4", "Std. Error"]
coef(summary(M4))["Var4", "z value"]
coef(summary(M4))["Var4", "Pr(>|z|)"]
logLik(M4)
rm(Var4)

```

# explore_model_for_Variant_5
rare, no association [2 of 100] vs [2 of 100]

```{r explore_model_for_Variant_5}

Var5 <- G[5,]
M5 <-glm(Y ~ Var5, family = binomial(link = "logit")) 
summary(M5)
coef(summary(M5))["Var5", "Estimate"]
coef(summary(M5))["Var5", "Std. Error"]
coef(summary(M5))["Var5", "z value"]
coef(summary(M5))["Var5", "Pr(>|z|)"]
logLik(M5)
rm(Var5)

```

# explore_model_for_Variant_6
common, no association [20 of 100] vs [20 of 100]

```{r explore_model_for_Variant_6}

Var6 <- G[6,]
M6 <-glm(Y ~ Var6, family = binomial(link = "logit")) 
summary(M6)
coef(summary(M6))["Var6", "Estimate"]
coef(summary(M6))["Var6", "Std. Error"]
coef(summary(M6))["Var6", "z value"]
coef(summary(M6))["Var6", "Pr(>|z|)"]
logLik(M6)
rm(Var6)

```

# explore_model_for_Variant7
near absolute association [99 of 100] vs [1 of 100]

```{r explore_model_for_absolute_association}

Var7 <- G[7,]
M7 <-glm(Y ~ Var7, family = binomial(link = "logit")) 
summary(M7)
coef(summary(M7))["Var7", "Estimate"]
coef(summary(M7))["Var7", "Std. Error"]
coef(summary(M7))["Var7", "z value"]
coef(summary(M7))["Var7", "Pr(>|z|)"]
logLik(M7)
rm(Var7)

```

# save_data

```{r save_data}

save.image(file="data/s22_simulation_to_test_variants.RData")

```

# final_section

```{r final_section}

sessionInfo()
Sys.time()

```

